Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models

### 2015-03-15 tex/parametrized-learning.tex
tweaks and decomposing test between mixture models

\begin{itemize}
 \item $x$: a vector of features
 \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
 \item $\theta$: parameters of a statistical model
\item $f(x| \theta)$:  probability density  (statistical model) for $x$ 
%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
%\item $f( s | \theta )$ The probability density function for $s$ implied by $f(x|\theta)$ and $s(x)$
\item $g( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $f(x|\theta_0, \theta_1)$ \end{itemize}
We will assume the $x_e$ are i.i.d., so that $f(D|\theta) = \prod_{e=1}^n f(x_e | \theta)$.

@@

Our main result is that one can form an equivalent test based on 
%\begin{equation}
%T'(D) = \prod_{e=1}^n \frac{ f(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ f(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
%\end{equation}
\begin{equation}\label{eq:equivLRtest}
T'(D) = \prod_{e=1}^n \frac{ g(s_e | \theta_0)}{ g(s_e | \theta_1)}
\end{equation}
if 
\begin{equation}
s_e = s(x_e; \theta_0, \theta_1) = \frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \; 
\end{equation}
or some monotonic function of that ratio. This will be proven below.
This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
used to learn $s(x; \theta_0, \theta_1)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$. In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the discriminative classifier explicitly parametrized in terms of $(\theta_0, \theta_1)$.

@@

Secondly, we are not interested in thresholded classification on individual events $x_e$, but are interested in a thresholded hypothesis test on an entire data set based on the likelihood ratio test statistic $T(D)$.  
Additionally, we know that both discriminative and generative classifier scores are often poorly calibrated. 
We wish to have well calibrated statements, but not about $p(y|x)$ but p-values associated to $P(T(D) > k |\theta)$.

Lastly, in the setting of frequentist hypothesis we do not have a prior $\pi(\theta)$. 
While we can use the generative models to produce training data $(x,y=0)$ generated 
from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$, the relative mix $p(y)$ 
is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then 
\begin{equation}
p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{f(x|\theta_1)}{f(x|\theta_0)+f(x|\theta_1))} \;,
\end{equation}
which is monotone with the desired likelihood ratio $f(x|\theta_1)/f(x|\theta_0)$ and free from the
complications of the prior $\pi(\theta)$ that is undefined in the frequentist context.

@@

The important part of the per-event function $q(x)$ is that it defines iso-contours in the feature space $x$. As we will show, our goal is to learn a monotonic function of $f(x|\theta_0)/f(x|\theta_1)$ that shares the same iso-contours. Then the remaining challenge is to find the appropriate rescaling that gives back  linear function of $q(x)$. Our claim is that the generative model $f(x|\theta)$ can be used to calibrate $g(s|\theta)$ and that
\begin{equation}
\ln T' = \sum_{e=1}^n \underbrace{\log \left[ \frac {g(s_e | \theta_0) }{ g(s_e | \theta_1) } \right]}_{q(s_e)} \;,
\end{equation}
leads to an equivalent test. In particular, we need to show the density
\begin{equation}
f(q_x|\theta) = \int dx \delta(q_x-q_x(x)) f(x|\theta)  / | \hat{n} \cdot \nabla q_x  | 
\end{equation}
is the same as
\begin{equation}
f(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, f(x|\theta) \, / | \hat{n} \cdot \nabla q_s  | \; .
\end{equation}
It is sufficient to show that $q(x_e) = q(s(x_e))$ $ \forall x\in\Omega_c$.

For notational simplicity, let $f_0(x) = f(x|\theta_0)$, $f_1(x) = f(x|\theta_1)$, and $s(x)=s(x; \theta_1, \theta_0)$.
The distribution of $x$ totally determines the distribution of $s$ via the change of variables $x\to s$. 
In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to $\Omega_c$ at the point $x$. The induced density $g_1(c)$ is given by 
\begin{equation}
g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x)  / | \hat{n} \cdot \nabla s  |
\end{equation}
and a similar equation for $g_0(c)$. 
%\textbf{Do we need Jacobian for x $\to$ s independent of delta function part, I think that's double counting?}

@@

Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{g}(x|\theta)$.

This treatment shows that in the asymptotic limit of large samples from the generative model, that we can approximate arbitrarily well the original likelihood ratio test. With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{g}(x)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test. This will depend in general on the ultimate goal of the test. We know that in the case of composite hypothesis tests that there is in general, not uniformly most powerful test, thus it is likely that a decision theoretic approach taking into account some weighting or utility over the space $\Theta$ is necessary. This is left as a subject for future work.

@@

In most settings that make use of likelihood ratio tests, the probability densities directly model the observed data via $f(x|\theta)$. In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$. In the parametrized setting this it is not possible to pre-compute $\hat s(x; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$, so we must computationally we must embed the classifier into the likelihood function to carry out the composition $g\circ s$. A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and discriminative classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}.

@@

There is an interesting distinction between this approach and the standard use in which the classifier is trained for a fixed $\nu_0$. In the standard use one trains a classifier for signal vs. background, which is equivalent (in an ideal setting) to training a classifier for  null (background-only) vs. alternate (signal-plus-backgound) as 
\begin{equation}
 \frac{f(x| 0, \nu_0)}{f(x|\hat \mu, \nu_0)} =  \frac{f_b(x| \nu_0)}{ \mu f_s( x_e \, |\,  \nu_0)  + (1-\mu)\, f_b( x_e \,|\, \nu_0)} = \left[ c_1 + c_2   \frac{f_s(x| \nu_0)}{ f_b( x_e \,|\, \nu_0)} \right ]^{-1} \; ,
\end{equation}
and $c_1$ and $c_2$ are constants. Specifically, the two approaches likelihood ratios are in one-to-one correspondence, so an ideal algorithm would lead to equivalent tests. In contrast, in the case of the generalized likelihood ratio test 
\begin{equation}
 \frac{f(x| 0, \hat{\hat{ \nu}})}{f(x|\hat \mu, \hat\nu)} =  \frac{f_b(x| \hat{\hat{ \nu}})}{ \hat \mu f_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, f_b( x_e \,|\, \hat \nu)} \; ,
\end{equation}
the background components don't cancel and there is an additional term $f_b(x| \hat{\hat{ \nu}})/f_b(x| {\hat{ \nu}})$.
In practice, with classifiers of finite capacity, there will be some tradeoff between taking into account this additional term and the more challenging learning problem when $\mu$ is very small.

>>>>

\begin{itemize}
 \item $x$: a vector of features
 \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
 \item $\theta$: parameters of a statistical model
\item $f(x| \theta)$:  probability density  (statistical model) for $x$ 
%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
%\item $f( s | \theta )$ The probability density function for $s$ implied by $f(x|\theta)$ and $s(x)$
\item $g( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $f(x|\theta)$ \end{itemize}
We will assume the $x_e$ are i.i.d., so that $f(D|\theta) = \prod_{e=1}^n f(x_e | \theta)$.

@@

Our main result is that one can form an equivalent test based on 
%\begin{equation}
%T'(D) = \prod_{e=1}^n \frac{ f(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ f(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
%\end{equation}
\begin{equation}\label{eq:equivLRtest}
T'(D) = \prod_{e=1}^n \frac{ g(s_e | \theta_0)}{ g(s_e | \theta_1)}
\end{equation}
if 
\begin{equation}
s_e = s(x_e; \theta_0, \theta_1) = M\left(\frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \right) \; 
\end{equation}
where $M$ is any monotonic function. This will be proven below.
This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
used to learn $s(x; \theta_0, \theta_1)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$. In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the discriminative classifier explicitly parametrized in terms of $(\theta_0, \theta_1)$.

@@

Secondly, we are not interested in thresholded classification on individual events $x_e$, but are interested in a thresholded hypothesis test on an entire data set based on the likelihood ratio test statistic $T(D)$.  
Additionally, we know that both discriminative and generative classifier scores are often poorly calibrated. 
We wish to have well calibrated statements regarding p-values defined by $P(T(D) > k |\theta)$, not well calibrated posterior probabilities $p(y|x)$.

Lastly, in the setting of frequentist hypothesis tests, we do not have a prior $\pi(\theta)$. 
While we can use the generative models to produce training data $(x,y=0)$ generated 
from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$, the relative mix $p(y)$ 
is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then 
\begin{equation}
p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{f(x|\theta_1)}{f(x|\theta_0)+f(x|\theta_1))} \;,
\end{equation}
which is monotone with the desired likelihood ratio $f(x|\theta_1)/f(x|\theta_0)$ and free from the
complications of the prior $\pi(\theta)$ that is undefined in the frequentist context.

@@

The important part of the per-event function $q(x)$ is that it defines iso-contours in the feature space $x$. As we will show, our goal is to learn a monotonic function of $f(x|\theta_0)/f(x|\theta_1)$ that shares the same iso-contours. Then the remaining challenge is to find the appropriate rescaling that gives back  linear function of $q(x)$. Our claim is that the generative model $f(x|\theta)$ can be used to calibrate $g(s|\theta)$ and that
\begin{equation}
\ln T' = \sum_{e=1}^n \underbrace{\log \left[ \frac {g(s_e | \theta_0) }{ g(s_e | \theta_1) } \right]}_{q(s_e)} \;,
\end{equation}
leads to an equivalent test. In particular, we need to show the density
\begin{equation}
f(q_x|\theta) = \int dx \delta(q_x-q_x(x)) f(x|\theta)  / | \hat{n} \cdot \nabla q_x  | 
\end{equation}
is the same as
\begin{equation}
f(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, f(x|\theta) \, / | \hat{n} \cdot \nabla q_s  | \; .
\end{equation}
It is sufficient to show that $q(x) = q(s(x))$ $ \forall x\in\Omega_c$.

For notational simplicity, let $f_0(x) = f(x|\theta_0)$, $f_1(x) = f(x|\theta_1)$, and $s(x)=s(x; \theta_1, \theta_0)$.
The distribution of $x$ totally determines the distribution of $s$. 
In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to $\Omega_c$ at the point $x$. The induced density $g_1(c)$ is given by 
\begin{equation}
g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x)  / | \hat{n} \cdot \nabla s  |
\end{equation}
and a similar equation for $g_0(c)$. 
%\textbf{Do we need Jacobian for x $\to$ s independent of delta function part, I think that's double counting?}

@@

Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{g}(\hat{s}|\theta)$.

This treatment shows that in the asymptotic limit of large samples from the generative model, that we can approximate arbitrarily well the original likelihood ratio test. With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{g}(\hat{s}|\theta)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test. This will depend in general on the ultimate goal of the test. We know that in the case of composite hypothesis tests that there is in general no uniformly most powerful test, thus it is likely that a decision theoretic approach taking into account some weighting or utility over the space $\Theta$ is necessary. This is left as a subject for future work.

@@

In most settings that make use of likelihood ratio tests, the probability densities directly model the observed data via $f(x|\theta)$. In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$. In the parametrized setting this it is not possible to pre-compute $\hat s(x; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$, so  we must embed the classifier into the likelihood function to carry out the composition $g\circ s$. A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and discriminative classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}.

@@

There is an interesting distinction between this approach and the standard use in which the classifier is trained for a fixed $\nu_0$. In the standard use one trains a classifier for signal vs. background, which is equivalent (in an ideal setting) to training a classifier for  null (background-only) vs. alternate (signal-plus-backgound) as 
\begin{equation}
 \frac{f(x| 0, \nu_0)}{f(x|\hat \mu, \nu_0)} =  \frac{f_b(x| \nu_0)}{ \mu f_s( x_e \, |\,  \nu_0)  + (1-\mu)\, f_b( x_e \,|\, \nu_0)} = \left[ c_1 + c_2   \frac{f_s(x| \nu_0)}{ f_b( x_e \,|\, \nu_0)} \right ]^{-1} \; ,
\end{equation}
and $c_1$ and $c_2$ are constants. Specifically, the two  likelihood ratios are in one-to-one correspondence, so an ideal algorithm would lead to equivalent tests. In contrast, in the case of the generalized likelihood ratio test 
\begin{equation}
 \frac{f(x| 0, \hat{\hat{ \nu}})}{f(x|\hat \mu, \hat\nu)} =  \frac{f_b(x| \hat{\hat{ \nu}})}{ \hat \mu f_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, f_b( x_e \,|\, \hat \nu)} \; ,
\end{equation}
the background components don't cancel and there is an additional term $f_b(x| \hat{\hat{ \nu}})/f_b(x| {\hat{ \nu}})$.
In practice, with classifiers of finite capacity, there will be some tradeoff between taking into account this additional term and the more challenging learning problem when $\mu$ is very small.

\subsection{Decomposing tests between mixture models into their components}

It is common that the generative model for the low-level features is a mixture model of several components
\begin{equation}
f(x|\theta)=\sum_c w_c(\theta) f_c(x| \theta) \;.
\end{equation}
In the case of particle physics, the distributions $f(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions. Moreover, when searching for a new particle, the null hypothesis would correspond to some of the coefficients $w_c=0$ while the alternate ``signal-plus-background'' hypothesis would have $0<w_{c\in \rm signal} \ll w_{c\in \rm background}$. In some cases $w_{c\in \rm signal} / w_{c\in \rm background} <10^{-6}$, which means the alternate hypothesis is a small perturbation to the null hypothesis. This can be a challenge for typical classifiers because they should devote their capacity to the region where $f_{c\in \rm signal}(x) / f_{c\in \rm background}(x)$ is relatively large. Lastly, even when the distributions $f_c(x|\theta)$ are well known, it is often the case that the coefficients are uncertain or treated as completely unknown. These all present challenges to machine learning algorithms that aim to learn $s(x; \theta_0, \theta_1)$.

%\begin{equation}
%\frac{f(x|\theta_0)}{f(x|\theta_1)}= \frac{\sum_c w_c(\theta_0) f_c(x| \theta_0)}{\sum_c w_c(\theta_1) f_c(x| \theta_1)} 
%\end{equation}
%
%\begin{equation}
%\frac{f(x|\theta_0)}{f(x|\theta_1)}= \sum_c \frac{ w_c(\theta_0) f_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)} 
%\end{equation}

However, it is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
\begin{eqnarray}
\frac{f(x|\theta_0)}{f(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) f_c(x| \theta_0)}{\sum_c w_c(\theta_1) f_c(x| \theta_1)} 
\\
%&=& \sum_c \frac{ w_c(\theta_0) f_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)} \\
%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)}{ w_c(\theta_0) f_c(x| \theta_0)}  \right]^{-1} \\
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ f_{c'}(x| \theta_1)}{  f_c(x| \theta_0)}  \right]^{-1} \\
&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ g_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ g_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
\end{eqnarray}
The second line is a trivial, but useful decomposition into pair-wise classification between $f_{c'}(x|\theta_1)$ and $f_c(x|\theta_0)$.  The third line uses the previous results to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$. In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $g_{c}(s_{c,c',\theta_0, \theta_1}| \theta_1)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
